\section{Tokenized-to-Tokenized Transformations}
\label{sec:symbolic_internal}

The transformations covered in this section share a common structural property: the source experience and the resulting carrier both reside in the tokenized layer, with no jump across the carrier spectrum. Within that layer the only axis of variation is the degree of formalization of the target carrier, and this axis matches the two tokenized sub-classes already defined in \S\ref{sec:carriers}---narrative tokenized and schematic tokenized. Section~\ref{sec:symbolic_internal:p1} covers the narrative-to-narrative case, in which raw natural-language trajectories are abstracted into more transferable but still narrative-form artifacts. Section~\ref{sec:symbolic_internal:p2} covers the narrative-to-schematic case, in which raw trajectories are formalized into graphs, workflows, or executable APIs whose reuse goes through parsing, execution, or graph traversal rather than through language understanding.

\subsection{Narrative-to-Narrative}
\label{sec:symbolic_internal:p1}

The simplest way to reuse a past trajectory is to keep it as raw text and prepend it---or a retrieved fragment of it---into the next prompt. This baseline works, but the literature consistently reports four costs of using it without modification: raw trajectories carry execution noise that competes with the actual decision signal, they occupy a substantial share of the context window, retrieval over them tends to surface surface-form matches rather than transferable structure, and the agent has no way to consolidate what is common across many trajectories of the same task family. The works grouped in this subsection share a single response to these costs: a language model is asked, without any weight update, to rewrite the raw trajectory into a shorter, less context-bound natural-language artifact that is then stored and retrieved on subsequent tasks. Across these works the underlying mechanism is uniform---an LLM-driven abstraction followed by retrieval and context-prepending---so they are not usefully distinguished by what they call their output (reflection, rule, summary, insight, skill, strategy, hint, and guideline are used interchangeably across papers and refer to overlapping artifacts). What they actually differ in is how the abstraction step itself is organized: how many trajectories enter one abstraction call, whether the abstraction is iterated, and whether more than one agent participates. The remainder of \S\ref{sec:symbolic_internal:p1} is organized along this axis, into four operation modes (\S\ref{sec:symbolic_internal:p1:mechanism}), followed by their multimodal realizations (\S\ref{sec:symbolic_internal:p1:multimodal}) and a unified discussion of their costs and characteristic weaknesses (\S\ref{sec:symbolic_internal:p1:discussion}).

\subsubsection{Mechanism}
\label{sec:symbolic_internal:p1:mechanism}

\paragraph{Single-Pass Extraction.}

In single-pass extraction, one trajectory is fed through one language-model call to produce one abstraction, with no iteration, no comparison against other trajectories, and no second agent involved. Reflexion is the canonical instance: at the end of each episode the agent produces a short verbal self-critique that is stored in an episodic buffer and prepended at the start of the next episode, so that improvement comes through linguistic feedback rather than gradient updates. Dynamic Cheatsheet maintains a self-curated note of strategies, code snippets, and problem-solving observations during inference; relevant fragments are retrieved and pasted into the prompt when a new question arrives, allowing the model to carry insights across questions without changing weights. RetroSum, introduced together with the AgentEHR clinical-interaction benchmark, performs a retrospective summarization at the end of each interaction so that long-context information loss in subsequent decisions is mitigated by a compact, regularly refreshed history. The structural bottleneck of this mode is that the quality of the abstraction is bounded by what a single LLM call can extract from one trajectory; once written, the artifact is treated as fixed, and there is no built-in mechanism for revising an early abstraction when later evidence contradicts it.

\paragraph{Iterative Refinement.}

Iterative refinement applies the abstraction operation repeatedly, with each pass taking the output of the previous pass as input rather than starting again from the raw trajectory. Iterative Experience Refinement makes the loop structure explicit and separates two patterns: a successive pattern that refines on top of the most recent experiences within a single task batch, and a cumulative pattern that consolidates across all previously seen batches. Training-Free GRPO uses a multi-round distillation loop to compress experiential knowledge into a token-level prior that takes effect at the moment of each subsequent API call, leaving the underlying weights untouched. AutoManual distributes the loop across three roles: a Planner produces actionable plans from the current rule set, a Builder updates the rule set online through a structured, case-conditioned prompting protocol after each new interaction, and a Formulator periodically compiles the accumulated rules into a single comprehensive manual; it is therefore an iterative-refinement system that also carries a cross-trajectory accumulation component. The structural bottleneck of this mode is that each refinement step takes the previous abstraction as its input instead of the original raw trajectory, so errors introduced early in the chain are not corrected later but compounded, and the artifact can drift into a self-consistent narrative that no longer reflects the underlying experience.

\paragraph{Multi-Trajectory Induction.}

Multi-trajectory induction aggregates several trajectories before producing a single abstraction, so the operation is many-to-one and surfaces patterns that no individual trajectory contains. ExpeL collects a body of interaction trajectories across a set of training tasks, extracts natural-language knowledge from that body, and at inference time recalls both the extracted insights and the supporting past experiences when responding to a new query. Trace2Skill dispatches a parallel fleet of sub-agents to analyze a diverse pool of executions, has each sub-agent extract trajectory-specific lessons, and then hierarchically consolidates the lessons through inductive reasoning into a single, conflict-free skill directory. ICE makes a three-stage pipeline explicit---it dynamically investigates planning and execution trajectories, consolidates them into simplified workflows and pipelines, and exploits those for subsequent task execution. The structural bottleneck of this mode is that the resulting abstraction is no better than the diversity of the trajectory pool it was induced from; failure modes and edge cases that the pool under-represents are precisely where the induced artifact will be brittle in deployment.

\paragraph{Multi-Agent / Multi-Role Generation.}

Multi-agent or multi-role generation distributes the abstraction operation across several agents or named roles that interact during the process; the determining factor of artifact quality is no longer the depth of any single call or iteration, but the dynamics among agents that hold different prompts, different information channels, or different objectives. MAR uses several debators with distinct personas to jointly produce a reflection, and reports that mutual critique among them mitigates the thought-degeneration commonly observed in single-agent self-reflection. AutoManual, already noted under iterative refinement, organizes its work as three named roles whose access to environment feedback differs: a Planner that consumes the rule set, a Builder that revises it after each interaction, and a Formulator that compiles long-form documentation. XSkill maintains two complementary streams---one giving concise action-level guidance for tool selection, the other giving structured task-level guidance for planning---both grounded in visual observations and jointly shaping downstream agent behavior. The structural bottleneck of this mode is that the outputs of different roles can disagree, and there is no built-in ground-truth arbitrator within the same system; introducing such an arbitrator turns the system into a composite that is no longer purely intra-narrative.

\paragraph{Artifact Diversity and Reuse.}

The artifacts produced across these four modes are referred to in the literature by many different names---reflection, rule, summary, insight, skill, strategy, guideline, hint, procedure---but these labels carry less analytical weight than they suggest: they overlap heavily in semantic content, and the differences between them are largely a matter of how the abstraction prompt was phrased rather than of any deeper distinction in the carrier. To preserve each paper's own choice of vocabulary without forcing it into a unified taxonomy, the prose makes no further effort to enumerate or sub-classify these labels. The way these artifacts are used at inference time is uniform across all four modes: a semantic-similarity retrieval over the stored artifacts selects the most relevant ones, which are then prepended into the current prompt for the language model to integrate during generation. As a consequence, whether a system based on this pathway is reliable in deployment depends jointly on two qualities---how faithful and how transferable the abstraction itself is, and how well the retrieval module surfaces the right artifact at the right moment---and improving either one yields measurable gains, so the two are properly viewed as parallel engineering levers rather than substitutes.


\subsubsection{Multimodal Realization}
\label{sec:symbolic_internal:p1:multimodal}

Among the multimodal instances of this pathway, the dominant pattern is one of asymmetric reduction: works take a multimodal source---GUI screenshot sequences, embodied interaction traces, vision-grounded conversations---and produce a textual artifact, while the inverse direction (a text-only source distilled into a multimodal artifact) is essentially absent. Agent~S processes GUI screenshots together with external knowledge search inside a hierarchical planning loop and stores the resulting reflections and summaries as text; ELITE extracts reusable strategies from visually grounded embodied trajectories and maintains them in an evolving textual strategy pool; GEMS keeps a hierarchical memory over multimodal interaction streams in which factual states are preserved verbatim while experiential observations are compressed into textual summaries. The asymmetry is structural rather than incidental, and the corresponding open space---text-source experience formalized into a multimodal artifact---is left to the open-problems section.

The single mechanical difference between multimodal and pure-text instances of this pathway lies one step earlier in the pipeline: the abstracting LLM operates on text, so any non-textual input must first be verbalized. Agent~S makes this step explicit by routing GUI screenshots through a dedicated Agent–Computer Interface module that translates them into structured textual descriptions before the reflection model is invoked. The cost of this step is that once the verbalization is wrong, the downstream abstraction has no further access to the original visual content, so a grounding error is silently inherited by every artifact derived from that trajectory.

A more fundamental difficulty in this pathway, when applied to multimodal experience, is not where to anchor an abstraction but what to abstract: different modalities of the same interaction are not redundant copies of one another but partial views that record different aspects, with vision capturing spatial layout and state change while text captures intent and causal structure, and neither modality being a strict superset of the other. An abstraction that is written in a single modality loses what only the other modality could express, and an abstraction that tries to carry several modalities at once pays the alignment cost on every read and write. Most current instances handle this by deliberately operating at episode-level granularity rather than at the level of individual perception–action pairs, treating the loss of fine-grained multimodal correspondence as an engineering trade-off they are willing to accept; finer-grained, frame-aligned multimodal abstraction remains an open direction.


\subsubsection{Discussion}
\label{sec:symbolic_internal:p1:discussion}

The entry cost of this pathway is low: human labels are not required because the language model itself plays the role of annotator, and the model used for generation can be the same one used for the abstraction step, so the system can be brought online with little upfront investment. Compute and storage cost stay modest under single-pass extraction, but they scale up under iterative refinement and under multi-agent setups, because every refinement round and every additional role implies extra inference calls and extra intermediate artifacts to keep around. The cost that does not amortize is the maintenance one: as the artifact store grows over time, deciding which entries have become stale or contradicted by the latest experience still requires a language model to make that judgment without an external ground truth, so the store tends to accumulate without principled eviction and the lifetime cost of running this pathway grows roughly with the lifetime of the system.

The gains delivered by this pathway are real but moderate, and they are best understood by comparison with the simpler baseline of retrieving raw trajectories and prepending them as in-context examples. Decision quality improves on tasks for which a similar past situation has been seen: WebCoach injects condensed cross-session advice into a web agent and noticeably stabilizes its behavior on previously visited domains, and RetroSum reduces long-context information loss in clinical decision-making by maintaining a regularly refreshed retrospective summary. Generalization improves because abstracting away instance-specific detail lets one piece of experience be retrieved by tasks that are structurally similar but verbally different, which is precisely where raw retrieval fails to fire; this gain still depends on retrieval-time semantic similarity, so it does not extend to genuinely far-transfer tasks. Inference efficiency is approximately neutral, with a modest advantage for this pathway: it avoids re-encoding a long raw trajectory on every call, but the abstraction itself still occupies part of the context budget. Taken together, the practical line between raw reuse and this pathway is the timescale over which experience accumulates---when a system is freshly deployed and the task distribution is narrow, raw retrieval is enough, but as the same task family is revisited many times the more compact, more retrievable abstractions begin to dominate, at the cost of having to maintain them from that point on.

A handful of weaknesses of this pathway recur across deployed systems, and they fall into two groups by where their root cause sits: the first three are tied to the LLM's self-judgment and are difficult to engineer away; the fourth concerns artifact management and yields straightforwardly to engineering. The most frequently reported weakness is that the abstraction can drift from the trajectory it claims to summarize---the language model writes a plausible-sounding causal attribution that no longer matches what the agent actually did, and the system has no external anchor to detect this; ExpeL's ablation, in which removing ground-truth task outcomes degrades reflection quality, is one indirect piece of evidence for this. Under iterative refinement the same drift compounds along the chain of refinements because each step reads the previous abstraction rather than the raw trajectory, and since this pathway does not require the raw trajectory to be retained the drift is often unrecoverable. Under multi-trajectory induction a complementary failure appears: when the trajectory pool fails to cover the relevant failure modes, a single successful instance or a narrow set of instances is generalized into a universal rule that breaks at the boundary; AutoGuide's choice to write each guideline together with the conditions under which it applies is one engineering mitigation, since it forces the artifact itself to record where it should and should not fire. The fourth weakness is the unbounded growth of the artifact store over time, which is observed in essentially every long-running deployment of this pathway; unlike the previous three it can be controlled by capping capacity, applying time-based eviction, or compressing clusters of similar artifacts into a single summary, and AutoAgent's Elastic Memory Orchestrator and the scaling characterization in ReasoningBank are representative engineering responses.

The artifacts produced along this pathway are utilized at inference time through Retrieve-and-Prepend (one of the six utilization modes catalogued in \S\ref{sec:cross_pathway:utilization:modes}).

\subsection{Narrative-to-Schematic}
\label{sec:symbolic_internal:p2}

The transformations covered in this subsection differ from those of the previous one in both representation and consumption, with the representational shift doing the main explanatory work: the target carrier is endowed with formal structural semantics---a graph topology, a workflow control flow, or a typed programmatic signature---rather than being left in free natural language. Once this structural semantics is in place, the question of whether a given input can use the artifact ceases to be one of language interpretation and becomes one of formal matching: an input that satisfies the schema is handled deterministically, an input that does not is rejected immediately. This single shift in the artifact's representation is what jointly explains the two characteristic properties that set this pathway apart from the narrative-to-narrative case---it offers repeatable, transferable behavior within the part of the input space the schema covers, and it fails sharply rather than gradually outside that part. The remainder of this subsection groups works by the kind of structure their artifacts carry: graphs, in which semantics live in the topology of nodes and edges; workflows, in which semantics live in explicit control flow; and code or APIs, in which semantics live in typed function signatures and executable bodies. This grouping correlates loosely with the strength of formalization but is not a single continuous spectrum.

\subsubsection{Mechanism}
\label{sec:symbolic_internal:p2:mechanism}

\paragraph{Graph.}

Graph-structured artifacts encode semantics into topology: nodes represent entities or events, edges represent relations, and reuse takes place through retrieval over the graph followed by traversal of the relevant subgraph. AriGraph constructs and incrementally updates a memory graph that integrates semantic and episodic memory while the agent explores its environment; Zep's core engine Graphiti is a temporally-aware knowledge graph that synthesizes unstructured conversational data and structured business data while preserving historical relations; SGMem represents dialogue as sentence-level graphs and uses them to capture associations across turn-, round-, and session-level contexts. GAAMA constructs a concept-mediated hierarchical knowledge graph through a three-step pipeline---verbatim episode preservation, LLM-based extraction of atomic facts and concepts, and higher-order reflection synthesis---that yields four node types and five edge types, with retrieval performed by an edge-type-aware Personalized PageRank scheme. CraniMem couples a neurocognitively motivated gated and bounded episodic buffer with a long-term knowledge graph, and runs a scheduled consolidation loop that replays high-utility traces into the graph while pruning low-utility entries. A-Mem occupies a boundary position within this group: its edges are drawn by semantic similarity between memory nodes, so retrieval over it behaves closer to ordinary similarity search than to genuine graph traversal, and the GAAMA ablation in which combining a graph-aware ranker with semantic search consistently outperforms pure semantic search is consistent with the more general observation that retrieval in this group is often a hybrid in practice.

\paragraph{Workflow / SOP / Pipeline.}

Workflow-class artifacts formalize an action sequence by making its control flow explicit---conditional branches, loops, parallel segments---so that the artifact itself can be read by an execution engine and stepped through. AFlow recasts workflow optimization as a search problem over code-represented workflows and uses Monte Carlo Tree Search to refine them through code modification, tree-structured experience, and execution feedback; A$^2$Flow extends this by introducing self-adaptive abstraction operators and an operator-memory mechanism so that earlier abstractions are reused across the search; Agent Workflow Memory induces commonly reused sub-routines from past traces and selectively injects the relevant workflow to guide subsequent generations. ProcMEM turns passive episodic narratives into executable skills defined by activation, execution, and termination conditions, trained with a Non-Parametric PPO scheme that combines semantic gradients for candidate generation with a PPO gate for verification, so its operational shape at reuse time---match activation, step through execution, check termination---is the shape of an execution engine rather than of a prepend-style memory. The script-abstraction branch of Mem$^p$ produces artifacts that are structurally isomorphic to those of Agent Workflow Memory and is best read as belonging to the same group.

\paragraph{Code / API / Programmatic.}

Code-class artifacts push formalization to its strongest form within this pathway by representing reuse as the invocation of an executable function. SkillWeaver lets a web agent autonomously explore a new website, practice candidate behaviors, and distill the resulting traces into reusable skills exposed as APIs for later agents to call. CoEvoSkills couples a Skill Generator with a Surrogate Verifier that co-evolves with it, so that the generator receives informative, actionable feedback during refinement without ever needing access to ground-truth test content. Meta Agent Search, presented as an instance of the Automated Design of Agentic Systems paradigm, has a meta-agent iteratively program new agents on top of an ever-growing archive of previous discoveries, and the agents it invents transfer across domains and across underlying models. Code as Policies sits at the lower-formalization end of this group: its code takes the form of high-level Python wrappers over abstract robot actions rather than low-level motor commands, so the artifact is closer to a thin programmatic surface over an existing action library than to a freshly authored function with full execution semantics.



\subsubsection{Multimodal Realization}
\label{sec:symbolic_internal:p2:multimodal}

When the source experience comes from a GUI agent, this pathway translates into programmatic artifacts with very little friction. The reason is structural: the action space of a GUI agent is intrinsically discrete---click, scroll, type, key-press, navigate---and these actions map almost directly to function calls with typed arguments, so no continuous-to-discrete bridging is needed. SkillWeaver exploits this alignment to consolidate web-agent interactions into callable APIs with very little custom schema engineering. In GUI settings the programmatic form is the default rather than one option among several.

The opposite picture appears in embodied continuous control. The raw signals collected there---motor commands, joint-angle trajectories, velocity profiles---do not match any natural discrete action set that could become the basis of typed function signatures, and discretizing them is a lossy operation in its own right. Code as Policies handles this by restricting its code to high-level action abstractions rather than low-level motor commands, which is a workable compromise but also reduces its formal strength relative to a fully programmatic artifact. When fine-grained motor experience genuinely needs to be reused, this pathway is no longer the right tool, and other carriers---latent compression or parametric internalization---become the relevant alternatives.

Putting the two halves of this picture together yields a three-tier gradient of how easily different modalities feed this pathway: text and GUI experience fit naturally because their action spaces are already discrete; web and code experience fit even more strongly because the interface itself is essentially already a programmatic artifact; embodied continuous control and audio sit at the difficult end because their underlying signals resist formalization without lossy discretization. This is not a strict continuous spectrum but the joint effect of two underlying factors---whether the action space is intrinsically discrete, and whether the interaction surface is already structured---and it is treated as such in the cross-pathway modality discussion later in the paper.

\subsubsection{Discussion}
\label{sec:symbolic_internal:p2:discussion}

The entry cost of this pathway sits in a different place than that of the previous one. There are no human labels to collect, but the schema must be designed up front: the node and edge types of a knowledge graph, the control primitives of a workflow, the typed signatures of an API library---these have to be fixed before any artifact can be produced. CoEvoSkills' typed skill schema and AriGraph's KG schema are representative of this front-loaded investment, which has no analog in the narrative-to-narrative case. The trade-off in return is that, once the schema stabilizes, the marginal cost of adding a new artifact is low and each new entry is structurally durable---it does not need to be re-evaluated against the latest task distribution to remain useful. The cost concentrates instead in schema evolution: when the original schema can no longer accommodate the kinds of tasks the system encounters, much of what has been accumulated under it may have to be reorganized.

The headline gain of this pathway is structural generalization: once a workflow has been formalized or an API has been defined, any input that satisfies the schema is routed through the same formal structure regardless of how textually similar it is to past inputs---SkillWeaver's discovered web-API skills, for instance, transfer across different sites in the same domain in a way that retrieval-based reuse of raw traces does not. The corresponding gain in the previous pathway is mediated by retrieval-time semantic similarity and is therefore probabilistic; here it is mediated by executable semantics and is therefore guaranteed within the schema's coverage. Set against this, the relevant baseline for cost comparison is not raw retrieval-then-prepend but a simpler intra-pathway variant in which the system relies on a few in-context structured examples without committing to a maintained schema. Compared to that variant, this pathway pays a higher up-front cost and a lower per-entry maintenance cost over time, and the point at which the two cost curves cross depends on how fast experience actually accumulates in the deployed setting---a property that drives the selection criterion taken up later in the paper.

The weaknesses of this pathway take a recognizably different shape from those of the previous one, though they can again be sorted into intrinsic and engineering kinds. The most direct intrinsic weakness is hard failure outside coverage: the formal grammar that gives the artifact its leverage also means that an input falling outside the schema produces a parse error or execution exception rather than the gradual quality degradation seen with narrative artifacts---useful in production because failures are visible and easy to diagnose, but inconvenient because graceful degradation is correspondingly hard to engineer. A more subtle intrinsic weakness is schema drift: the syntax of a node, edge, or function may remain valid while the real-world entity or behavior it refers to has quietly moved on, and the system continues to operate with a kind of false confidence whose detection requires external grounding. In the workflow group, search-based discovery has its own bias: under limited rollout budgets MCTS tends to converge prematurely to local optima, which AFlow and A$^2$Flow both reflect through run-to-run performance variance, and the search space grows exponentially with workflow depth, sharpening the problem as workflows get longer. Finally, in the graph group, long-running deployments tend to develop a kind of staleness that has no parallel in the previous pathway: correct entries are silently overwritten by incorrect ones, which is a correctness problem rather than the volume problem of unbounded growth, and it implies that even a stable schema does not by itself guarantee the truthfulness of what is stored under it.

Artifacts produced along this pathway are utilized at inference time through Retrieve-and-Execute for the workflow and code groups and through Retrieve-and-Traverse for the graph group, both catalogued in the utilization taxonomy of \S\ref{sec:cross_pathway:utilization:modes}.

Both groups of works collected in this section keep the resulting artifact in the tokenized layer; what separates them is whether that artifact is also given a formal structure. When it is, the way the artifact is used and the way its cost accumulates over time both change, but the underlying skeleton---abstraction by a language model without weight updates, followed by retrieval-mediated reuse---is shared across them. Section~\ref{sec:symbolic_to_latent} takes the first step out of this layer, by examining how tokenized experience can be compressed into a latent representation and what is gained and lost in the move.
